---
title: "Job Market Hub"
page-layout: full
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(lubridate)
library(googlesheets4)
library(kableExtra)
library(fontawesome)

## package global options
# skip googlesheets login
gs4_deauth()

## Set date threshold to display jobs
t = "2026-10-31"
```

## Upcoming Deadlines

```{r, message = FALSE}
jobs = read_sheet("https://docs.google.com/spreadsheets/d/1gEXP5EvvPtKmxY0srki6cNaa3JxMslzdHmfzrLB-PGM/edit?gid=1792307319#gid=1792307319",
                  sheet = 1)

# some manipulation to show only what is useful
jobs = jobs %>% 
# don't show if already applied
  filter(is.na(applied)) %>% 
# filter deadlines within treshold
filter(as_date(deadline) <= as_date(t)) %>% 
  mutate(
  # url = paste0('<a href="', url,'"target="_blank"></a>'),
  deadline = format(deadline, "%B %e, %Y")
)
```

**Last update:** `r format(Sys.Date(), "%B %e, %Y")`. Full list [here](https://docs.google.com/spreadsheets/d/1gEXP5EvvPtKmxY0srki6cNaa3JxMslzdHmfzrLB-PGM/edit?gid=0#gid=0).

**Number of letters to write:**

```{r}
jobs %>% 
  summarize(across(Jake:Michelle, ~ sum(.x, na.rm = TRUE))) %>% 
  select_if(~sum(.) > 0) %>% 
  kbl() %>% 
  kable_paper(full_width = FALSE, position = "left")
```

&nbsp;

<div id="writer-filter">
<strong>Filter by letter writer:</strong>&nbsp;
<button class="filter-btn active" data-writer="all">All</button>
<button class="filter-btn" data-writer="jake">Jake</button>
<button class="filter-btn" data-writer="matt">Matt</button>
<button class="filter-btn" data-writer="michelle">Michelle</button>
<span id="filter-count"></span>
</div>

&nbsp;

```{r, results = "asis"}
refs = jobs %>% select(Jake:Michelle)

for(i in 1:nrow(jobs)){
  writers_for_job = paste(tolower(colnames(refs)[which(refs[i,] == 1)]), collapse = " ")
  title_html = paste0('<a href="', jobs$url[i], '"><strong>',
                      jobs$position[i], ' - ', jobs$department[i], ' - ', jobs$place[i],
                      '</strong></a>')
  letters_str = paste(colnames(refs)[which(refs[i,] == 1)], collapse = ", ")
  cat(paste0(
    '<div class="job-entry" data-writers="', writers_for_job, '">\n',
    title_html, '<br/>\n',
    '<strong>Letters:</strong> ', letters_str, '<br/>\n',
    '<strong>Deadline:</strong> ', jobs$deadline[i], '<br/>\n',
    '<strong>Delivery:</strong> ', jobs$letters[i], '<br/>\n',
    '<strong>Notes:</strong> ', jobs$notes[i],
    '\n</div>\n\n'
  ))
}
```

<script>
document.addEventListener('DOMContentLoaded', function () {
  const buttons = document.querySelectorAll('.filter-btn');
  const entries = document.querySelectorAll('.job-entry');
  const countEl = document.getElementById('filter-count');

  function applyFilter(writer) {
    let visible = 0;
    entries.forEach(entry => {
      const writers = entry.dataset.writers || '';
      const show = writer === 'all' || writers.split(' ').includes(writer);
      entry.style.display = show ? '' : 'none';
      if (show) visible++;
    });
    buttons.forEach(btn => btn.classList.toggle('active', btn.dataset.writer === writer));
    countEl.textContent = writer !== 'all'
      ? `Showing ${visible} job${visible !== 1 ? 's' : ''} for ${writer.charAt(0).toUpperCase() + writer.slice(1)}`
      : '';
    const url = new URL(window.location);
    writer === 'all' ? url.searchParams.delete('writer') : url.searchParams.set('writer', writer);
    history.replaceState({}, '', url);
  }

  buttons.forEach(btn => btn.addEventListener('click', () => applyFilter(btn.dataset.writer)));

  const param = new URLSearchParams(window.location.search).get('writer');
  if (param) applyFilter(param.toLowerCase());
});
</script>


## Materials

- [Short CV](diaz_cv.pdf) (For job applications)
- [Long CV](cv.pdf)
- [Methods cover letter](cover.pdf)
- [Methods research statement](research.pdf)
- [CP cover letter](cover_cp.pdf)
- [CP research statement](research_cp.pdf)
- [Teaching statement](teaching.pdf)
